---
title: "cm009"
author: "Anita"
date: "October 2, 2018"
output: 
  html_document: 
    keep_md: yes
---

---
title: "cm009 Exercises: tidy data"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---


when you are working in R, R knows where it is working from on your computer, accessing a folder
to figure out where R is working from you can use getwd()
can also setwd()


whenever you have a character vector, base R will load it as a factor which causes problems
but tidyverse R will keep it as character factor (instead of treating it like it's all integers)


tibbles are data frames with soem extra features
base R way of making data frame is with data.frame function, with tibble:  tibble()
```{r}
suppressPackageStartupMessages(library(tidyverse))
```

## Reading and Writing Data: Exercises

Make a tibble of letters, their order in the alphabet, and then a pasting of the two columns together.

tibble backwards compatble, don't need to first define let and num columns, tibble functions will make as you go along (just 1 function for several columns)

so here we cretae a frame with the columns letters and numbers and then put them together

letters is a vector in R, if it wasn't we could have added one manually from a:z
```{r}
tibble(let = letters,
       num = 1:length(letters),
       comb = paste(let, num))

tibble(let = letters,
       num = 1:length(letters),
       comb = paste0(let, num))
```

Make a tibble of three names and commute times.


~ indicates that these are the names we will create in data frame
```{r}
tribble(
  ~name, ~time,
  "Frank", 30,
  "Lisa", 10,
  "Mark", 50
)
```


Write the `iris` data frame as a `csv`. 

```{r}

write_csv(iris, "")# note if I click tab inside quotation marke everything in my current working directory comes up to choose from

write_csv(iris, "iris.csv")
```

Write the `iris` data frame to a file delimited by a dollar sign. 

```{r}
write_delim(iris, "iris.txt", delim = "$")

```

Read the dollar-delimited `iris` data to a tibble.

```{r}
read_delim("iris.txt", delim = "$")

```

Read these three LOTR csv's, saving them to `lotr1`, `lotr2`, and `lotr3`:

- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv
- https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv
- https://github.com/jennybc/lotr-tidy/blob/master/data/The_Return_Of_The_King.csv

```{r}
lotr1 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Fellowship_Of_The_Ring.csv")
lotr2 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Two_Towers.csv")
lotr3 <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/The_Return_Of_The_King.csv")
```




tidy is long form data, and untidy is in wide form

I should check lotr1, lotr2, and lotr3 and make sure that the data frames are the same if I want t stack them on top of each other (combine them)
i.e., same types of variables with same names etc. if this is not the case I first have to do some data wrangling


## `gather()`

(Exercises largely based off of Jenny Bryan's [gather tutorial](https://github.com/jennybc/lotr-tidy/blob/master/02-gather.md))

This function is useful for making untidy data tidy (so that computers can more easily crunch the numbers).

1. Combine the three LOTR untidy tables (`lotr1`, `lotr2`, `lotr3`) to a single untidy table by stacking them.  

```{r}
(lotr_untidy <- bind_rows(lotr1, lotr2, lotr3))
#looks like it is tidy bt it is not because we ant to combine male and female in one column.
#this is what long forma means!
```

2. Convert to tidy. Also try this by specifying columns as a range, and with the `contains()` function.

```{r}
gather(lotr_untidy, key="Gender", value="Words", Female, Male)#that will name the new columns
#and the Female, Male means that this is what the column will be made of (spred across its own column)
#Female, and Male are the specified columns that I want to be put in this new column

#key: new line I want to make called gender
#value: what s in the line, here the word count, i.e., how much words did each individual speak
#and then behind this I specify which columns of the wide form go in there





gather(lotr_untidy, key="Gender", value="Words", Female:Male) 
#using colons means that I choose the range of columns from Female to Male
gather(lotr_untidy, key="Gender", value="Words", contains("ale"))
#include all columns where the name of the column contains the letters "ale"
```

3. Try again (bind and tidy the three untidy data frames), but without knowing how many tables there are originally. 
    - The additional work here does not require any additional tools from the tidyverse, but instead uses a `do.call` from base R -- a useful tool in data analysis when the number of "items" is variable/unknown, or quite large. 

```{r}
lotr_list <- list(lotr1, lotr2, lotr3)
lotr_list

#now if we want to bind all of these together, we can se the do.call function
do.call(bind_rows, lotr_list)

#then from here we can mae it tidy again

```

## `spread()`

(Exercises largely based off of Jenny Bryan's [spread tutorial](https://github.com/jennybc/lotr-tidy/blob/master/03-spread.md))

This function is useful for making tidy data untidy (to be more pleasing to the eye).

Read in the tidy LOTR data (despite having just made it):

```{r}
lotr_tidy <- read_csv("https://raw.githubusercontent.com/jennybc/lotr-tidy/master/data/lotr_tidy.csv")
```

Get word counts across "Race". Then try "Gender". 

```{r}
spread(lotr_tidy, key = "Race", value = "Words")
```

Now try combining race and gender. Use `unite()` from `tidyr` instead of `paste()`. 

```{r}
lotr_tidy %>%
  unite(Race_Gender, Race, Gender) %>%
  spread(key = "Race_Gender", value = "Words")
#unite function: first command is the name of the new column



lotr_tidy %>%
  mutate(x = rnorm(nrow(lotr_tidy))) %>%
   spread(key = "Gender", value = "x")
#x is a randomly generated variable

```

## Other `tidyr` goodies

Check out the Examples in the documentation to explore the following.

`expand` vs `complete` (trim vs keep everything). Together with `nesting`. Check out the Examples in the `expand` documentation.

```{r}
?expand
head(mtcars)
expand(mtcars, vs, cyl)

#returns a combination of all possible combnations in teh data frame even if it doesn't actually exist in the data frame


df <- tibble(
  year   = c(2010, 2010, 2010, 2010, 2012, 2012, 2012),
  qtr    = c(   1,    2,    3,    4,    1,    2,    3),
  return = rnorm(7)
)
df

df %>% expand(year, qtr)


df %>% expand(year = full_seq(year, 1), qtr)
df %>% complete(year = full_seq(year, 1), qtr)
#shows every year even if there is no data



experiment <- tibble(
  name = rep(c("Alex", "Robert", "Sam"), c(3, 2, 1)),
  trt  = rep(c("a", "b", "a"), c(3, 2, 1)),
  rep = c(1, 2, 3, 1, 2, 1),
  measurment_1 = runif(6),
  measurment_2 = runif(6)
)
experiment
#goal is to fill in the stuff that isn't there yet with NA

all <- experiment %>% expand(nesting(name, trt), rep)
all
```

`separate_rows`: useful when you have a variable number of entries in a "cell".

```{r}
?separate_rows

df <- data.frame(
  x = 1:3,
  y = c("a", "d,e,f", "g,h"),
  z = c("1", "2,3,4", "5,6"),
  stringsAsFactors = FALSE
)
df

#best way to make it tidy is to enter a new row for each cell

separate_rows(df, y, z, convert = TRUE)
```

`unite` and `separate`.

```{r}
?unite
?separate
```

`uncount` (as the opposite of `dplyr::count()`)

```{r}
?uncount
df <- tibble::tibble(x = c("a", "b"), n = c(1, 2))
df

uncount(df, n)
```


`drop_na` and `replace_na`
drop rows with na or replace na with some values
```{r}


```

`fill`

```{r}

```


`full_seq`

```{r}

```


## Time remaining?

Time permitting, do [this exercise](https://github.com/jennybc/lotr-tidy/blob/master/02-gather.md#exercises) to practice tidying data. 