---
title: 'cm005: `dplyr` Exercise'
output: 
  html_document:
    keep_md: yes
---

**Optional, but recommended startup**:

Change the file output to both html and md _documents_ (not notebook).

# Intro to `dplyr` syntax

1. Load the `gapminder` and `tidyverse` packages. Hint: `suppressPackageStartupMessages()`!
    - This loads `dplyr`, too.
2. `knit` the document. 

```{r}
library(gapminder)
library(tidyverse)
```


## `select()`

1. Make a data frame containing the columns `year`, `lifeExp`, `country` from the gapminder data, in that order.

Ctrl Alt I to create new coding chunk
```{r}
select(gapminder, year, lifeExp, country)
head(gapminder)
```

2. Select all variables, from `country` to `lifeExp`. 
```{r}
select(gapminder, country:lifeExp)
head(gapminder)
```



3. Select all variables, except `lifeExp`.
```{r}
select(gapminder, -lifeExp)
```
if you don't capture the output of the function, everything stays as is
select is only returning the modified output but it's not stored anywhere
to store:
```{r}
a<-select(gapminder, -lifeExp)
head(a)
```


4. Put `continent` first. Hint: use the `everything()` function.
```{r}
select(gapminder, continent, everything())
```

5. Rename `continent` to `cont`.
```{r}
rename(gapminder, cont=continent)
#to store:
  b<-rename(gapminder, cont=continent)
  b
```

## `arrange()`

1. Order by year.
```{r}
arrange(gapminder, lifeExp)#default: ascending
arrange(gapminder, desc(lifeExp))
```

2. Order by year, in descending order.
```{r}
arrange(gapminder, desc(lifeExp))
```

3. Order by year, then by life expectancy.
```{r}
arrange(gapminder, year, lifeExp)#first ordered by year and then by life Exp., goes down the list
```


## Piping, `%>%`

Note: think of `%>%` as the word "then"!

1. Combine `select()` Task 1 with `arrange()` Task 3.

```{r}
gapminder %>%
  select(year, lifeExp, country)%>% #no need to call gapminder again, taes from previous function
  arrange(year, lifeExp)
#1. took gapminder data set
#2. select 3 variables
#3. arrange in a certain order
#tidyverse strives for a self-documenting code; i.e., the code is self-explanatory

#to store it all:
b<-gapminder %>%
  select(year, lifeExp, country)%>% #no need to call gapminder again, taes from previous function
  arrange(year, lifeExp)


#by default whenever you use ipng, the output of the first function will be pt in the first place 
#i.e., data of the next function, so here in the select function it really says gapminder first
#way to avoid
#f(a=3, b=.)
```


## `filter()`

1. Only take data with population greater than 100 million.
```{r}
gapminder%>%
  filter(pop>100000000)

gapminder%>%
  filter(!(pop>100000000))#this now menas NOT >100000000, so same as <=

```

2. Of those, only take data from Asia.

## git stuff (Optional)

Knit, commit, push!

# Break/Challenge: metaprogramming

Here's an activity for you to do during the break, in case you're all caught up. It should help you understand metaprogramming a bit more.

Suppose you're the instructor of an R programming course. You write an assignment question to evaluate whether students can write an `if` statement, for which an answer to the question looks something like this:

```
my_commute <- 60
if (my_commute > 30) {
    print("That's a long commute!")
} else {
    print("That's a short commute.")
}
```

Your task is to use metaprogramming to check whether a response (like the one above) works and contains an `if` statement. You should roughly follow these steps, using [adv-r: expressions](https://adv-r.hadley.nz/expressions.html) as a resource (especially Section 18.1).

1. Wrap the above block of code in the `expr()` function from the `rlang` package.
2. Use the `eval()` function to execute the code, to see if the code runs.
3. Use the `as.character()` function to check whether this response contains an `if` statement.

# Relational/Comparison and Logical Operators in R

1. Find all entries of Canada and Algeria occuring in the '60s. 

```{r}
gapminder%>%
  filter((country=="Canada" |country=="Algeria") & year >=1960 & year < 1970)


gapminder%>%
  filter(country %in% c("Canada", "Algeria"),
         year >=1960,
          year < 1970)
```




2. Find all entries of Canada, and entries of Algeria occuring in the '60s. 
3. Find all entries _not_ including Canada and Algeria.

# Bonus Exercises

If there's time remaining, we'll practice with these three exercises. I'll give you 1 minute for each, then we'll go over the answer.

1. Take all countries in Europe that have a GDP per capita greater than 10000, and select all variables except `gdpPercap`. (Hint: use `-`).

2. Take the first three columns, and extract the names.

3. Of the `iris` data frame, take all columns that start with the word "Petal". 
    - Hint: take a look at the "Select helpers" documentation by running the following code: `?tidyselect::select_helpers`.
    - Exercise from [r-exercises](https://www.r-exercises.com/2017/10/19/dplyr-basic-functions-exercises/).
