---
title: "cm011 Exercises: R as a programming language"
output: html_document
---

# Part I

## Types and Modes and Classes, Oh My!

R objects have a _type_, a _mode_, and a _class_. This can be confusing:

```{r}
a <- 3
print(typeof(a)) # double: how much decimals can be stored
print(mode(a))  #numeric
print(class(a))
```


```{r}
print(typeof(iris)) #list
print(mode(iris))  #list
print(class(iris))
```


```{r}
print(typeof(sum)) #built in to R
print(mode(sum)) #function
print(class(sum))
```

Usually, there's no need to fuss about these differences: just use the `is.*()` family of functions. Give it a try:

```{r}
is.numeric(a)  
is.data.frame(iris)

#don't really need to worry about what type.class etc. something is, just check if it is what I want it to be
```

We can also coerce objects to take on a different form, typically using the `as.*()` family of functions. We can't always coerce! You'll get a sense of this over time, but try:

- Coercing a number to a character.
- Coercing a character to a number.
- Coercing a number to a data.frame. `letters` to a data.frame.

```{r}
as.character(100)
as.numeric("100")
as.numeric("100") * 2
#as.numeric("hello") # doesn't work!
as.data.frame(letters)
```

There is also a slight difference between coercion and conversion, but this is usually not important.

## Vectors

Vectors store multiple entries of a data type. You'll discover that they show up just about everywhere in R, so they're fundamental and extremely important.

### Vector Construction and Basic Subsetting

We've seen vectors as columns of data frames:

```{r}
mtcars$hp
```

Use the `c()` function to make a vector consisting of the course code (`"STAT"` and `545`). Notice the coercion. Vectors must be homogeneous.

```{r}
(course <- c("STAT", 545)) # a vector can only hold one type of information (so here characters, so 545 is now treated as a character)
```


Subset the first entry. Remove the first entry. Note the base-1 system.

```{r}
course[1] #gives you first entry only
course[-1]
course
sort(course)[1]
```

Use `<-` to change the second entry to "545A". Using the same approach, add a third entry, "S01".

```{r}
course[2] <- "545A"

"545A" <- course[2] #is there a difference?

course[3] <- "s01"
course
```

Subset the first and third entry. Order matters! Subset the third and first entry.

```{r}
course[c(3,1)]

```

Explore integer sequences, especially negatives and directions. Especially `1:0` that might show up in loops!

```{r}
3:10
10:-5

1:0

#Vector of length zero:
seq_len(0)
seq_len(10)
```

Singletons are also vectors. Check using `is.vector`.

```{r}
is.vector(6)
a[1][1]


c("hi", "you") # a vector of 2 characters
paste("hi", "you") #make one character out of it

v1 <- "hi"
v2 <- "you"
paste (v1, v2) #after creating two vectors, put them toether in a matrix

```


### Vectorization and Recycling

A key aspect of R is its vectorization. Let's work with the vector following vector:

```{r}
(a <- 7:-2)
(n <- length(a))
```

Square each component:

```{r}
a ^ 2  #R does this to every component, don't have to specify for each 
```

Multiply each component by 1 through its length:

multiply first number if 1, next with 2, then with 3 etc.

```{r}
a * 1:10 #component wise operation
```

It's important to know that R will silently recycle! Unless the length of one vector is not divisible by the other. Let's see:

```{r}
a * 1:3
#R will take the shorter vector of 1, 2, 3 and will duplicate it until it matches the length of the other vector
#thus it multiplies the fiirst number with 1, the second with 2, the third with 3, the fourth with 1 etc. etc.
#that's called recycling
#this is a problem if I don't intend it to do that

a*1:2
#this line won't give a warning because the shorter vector isn't choped off (here every even number is multiplied by 2)

```


This is true of comparison operators, too. Make a vector of logicals using a comparison operator.

```{r}
a > 0 # this is a logical vector of TRUEs and FALSEs
```


Now try a boolean operator. Note that && and || are NOT vectorized!

```{r}
a > 0 & a < 5 #returns TRUE only wherever both are true

a > 0 && a < 5 #this is NOT vectorized, may see this in if statemnts because we just want either TRUE or False, it only looks at the first entry of the left hand vector and the first entry of the right hand vector
#but this makes little sense in an if statement if it is only based on first entry in vector
#see below for solution

#Use functions "all or "any to reduce a vector of non-1 length
any(a > 0 & a <5)
all(a > 0 & a <5)

a > 0 || a < 5
```


Recycling works with assignment, too. Replace the entire vector a with 1:2 repeated:

```{r}
a[1:n] <- 1:2 #subset first entry all way to last entry 
a 
```


### Special Subsetting

We can subset vectors by names and logicals, too.

Recall the course vector:

```{r}
course <- c("STAT", "545A", "S01")
```

Let's give the components some names ("subject", "code", and "section") using three methods:

1. Using the setNames function.
- Notice that the vector does not change!!

```{r}
#useful for piping!
setNames(course, c("subject", "code", "section"))
```


2. Using the names function with `<-`. Also, just explore the names function.

```{r}
names(course) <- c("subject", "code", "section")
course
names(course)
```

3. Re-constructing the vector, specifying names within `c()`.

```{r}
course <- c(subject="STAT", code="545A", section="S01")
```


Subset the entry labelled "section" and "subject". 

```{r}
course["section"]
course[c("section", "subject")]
course[["section"]]
```

Amazingly, we can also subset by a vector of logicals (which will be recycled!). Let's work with our integer sequence vector again:

```{r}
(a <- 7:-2)
(n <- length(a))

a[a > 0] #omits all entries that are less than or equal to zero
```

## Lists

Unlike vectors, which are atomic/homogeneous, a list in R is heterogeneous. 
If a vector has different types it will transform everything in the least informative one (usually character)

Try storing the course code (`"STAT"` and `545`) again, but this time in a list. Use the `list()` function.

```{r}
(course <- list("STAT", 545))
```


Lists can hold pretty much anything, and can also be named. Let's use the following list:

```{r}
(my_list <- list(year=2018, instructor=c("Vincenzo", "Coia"), fav_fun=typeof))

#you can store anything in a list (a vector of length 1, a vector of length 2, and a functin are stored in this list)
```


Subsetting a list works similarly to vectors. Try subsetting the first element of `my_list`; try subsettig the first _component_ of the list. Notice the difference!

```{r}
my_list[1]
my_list[[1]]

my_list[2]
my_list[[2]]
```

Try also subsetting by name:

```{r}
my_list["year"]
my_list[["year"]]

my_list$year
```

Smells a little like `data.frame`s. It turns out a `data.frame` is a special type of list:

```{r}
(small_df <- tibble::tibble(x=1:5, y=letters[1:5]))
```

```{r}
is.list(small_df)
```

```{r}
as.list(small_df)

diag(5)
#Matrix can't hold objects of different types whereas data fraes can (e.g., characters and integers)
```

Note that there's a difference between a list of one object, and that object itself! This is different from vectors.

```{r}
identical(list(4), 4) #here we are looking at a list, lists have extra baggage around them
identical(c(4), 4) #here we are looking at a vector
```

# Part II

## Global Environment

When you assign variables in R, the variable name and contents are stored in an R environment called a global environment.

See what's in the Global Environment by:

- Executing `ls()`.
- Looking in RStudio, in the "Environments" pane.

```{r}

```


Making an assignment "binds" an object to a name within an environment. For example, writing `a <- 5` assigns the object `5` to the name `a` in the global environment.

The act of "searching for the right object to return" is called scoping.

By the way: the global environment is an object, too! It's the output of `globalenv()`, and is also stored in the variable `.GlobalEnv`:

```{r}
globalenv()
.GlobalEnv
```

## The Search Path

How does R know what `iris` is, yet `iris` does not appear in the global environment? What about functions like `length`, `sum`, and `print` (which are all objects, too)?

Let's explore.

1. Each package has its own environment.
    - Install and load the `pryr` package, and use `ls()` to list its bindings (its name is "package:pryr").

```{r}

```

2. There's a difference between an _environment_ and its _name_. Get the environment with name "package:pryr" using the `as.environment()` function.

```{r}

```


3. Each environment has a parent. Use `parent.env()` to find the parent of the global environment.

```{r}

```


4. There are packages that come pre-loaded with R, and they're loaded in a sequence called the search path. Use `search()` to identify that path; then see it in RStudio.

```{r}

```

First scoping rule: R looks to the parent environment if it can't find an object where it is.

5. Use `pryr::where()` to determine where the first binding to the name `iris` is located.

```{r}

```


6. Override `iris` with, say, a numeric. Now `where()` is it? Can you extract the original?

```{r}

```


7. Override `sum` with, say, a numeric. `where()` is `sum` now? Can you still use the original `sum()` function?

```{r}

```

Special scoping rule for functions! R knows whether or not to look for a function.

8. Look at the source code of the `pryr:where()` function. It contains a line that creates a binding for "env". Why, then, is `env` nowhere to be found? Answer: execution environments. 

```{r}
#pryr::where
```



